# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

```{r}
data <- read.csv("activity.csv")
data2 <- data[complete.cases(data), 1:2]
library(reshape2)
melted_data <- melt(data2, id.vars="date")
tdset <- dcast(melted_data, date ~ variable, sum)
summary(tdset)
```

## What is mean total number of steps taken per day?

```{r}
barplot(tdset$steps, names.arg = tdset$date, xlab = "date", ylab = "steps")
```
### mean number of steps taken per day
```{r}
mean(tdset$steps)
```
### median number of steps taken per day
```{r}
quantile(tdset$steps, probs=0.5)
```
## What is the average daily activity pattern?
```{r}
data3 <- data[complete.cases(data), c("steps", "interval")]
melted_data2 <- melt(data3, id.vars="interval")
tdset2 <- dcast(melted_data2, interval ~ variable, mean)
plot(tdset2, type = "l")
```
### Which 5-minute interval contains the maximum number of steps?
```{r}
tdset2[which(tdset2$steps==max(tdset2$steps)),]$interval
```

## Imputing missing values
### total number of missing values in the dataset
```{r}
sum(!complete.cases(data))
```
### filling missing NAs by the previously calculated mean
```{r}
newdata <- data
newdata[is.na(newdata)] <- mean(tdset2$steps)
melted_data3 <- melt(newdata[,1:2], id.vars="date")
tdset3 <- dcast(melted_data3, date ~ variable, sum)
head(tdset3)
barplot(tdset3$steps, names.arg = tdset3$date, xlab = "date", ylab = "steps")
```
### mean number of steps taken per day after filling missing NAs
```{r}
mean(tdset3$steps)
```
### median number of steps taken per day after filling missing NAs
```{r}
quantile(tdset3$steps, probs=0.5)
```
The mean does not differ at all from the previous value but median does differ a little bit. So the impact of imputing missing data on the estimates of the total daily number of steps is negligibly small.

## Are there differences in activity patterns between weekdays and weekends?
### Creates a new factor variable with two levels – “weekday” and “weekend” 
```{r}
daysID <- function(date) {
    if (weekdays(as.Date(date)) %in% c("Saturday", "Sunday")  ) {
        "weekend"
    } 
    else {
        "weekday"
    }
}
newdata$daysID <- as.factor(sapply(newdata$date, daysID))
tail(newdata)
```
```{r}
### Panel plot containing a time series plot of the 5-minute interval 
par(mfrow = c(2, 1))
for (type in c("weekend", "weekday")) {
    steps.type <- aggregate(steps ~ interval, data = newdata, subset = newdata$daysID == 
        type, FUN = mean)
    plot(steps.type, type = "l", main = type)
}
```