# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data

```{r echo=TRUE}
data <- read.csv("activity.csv")
library(reshape2)
melted_data <- melt(data, na.rm = TRUE, id.vars="date")
tdset <- dcast(melted_data, date ~ variable, sum)
summary(tdset)
```

## What is mean total number of steps taken per day?

```{r}
library("ggplot2")
p <- ggplot(tdset, aes(x=steps)) + geom_histogram(colour="white")
p + labs(title='Histogram of number of steps taken daily\n',
       y='Count', x='Steps')
```
### mean number of steps taken per day
```{r}
mean(tdset$steps)
```
### median number of steps taken per day
```{r}
quantile(tdset$steps, probs=0.5)
```
## What is the average daily activity pattern?
```{r}
melted_data2 <- melt(data[, c("steps", "interval")], na.rm = TRUE, id.vars="interval")
tdset2 <- dcast(melted_data2, interval ~ variable, mean)
p1 <- ggplot(tdset2, aes(x=interval, y=steps)) + geom_line()
p1 + labs(title='Time series plot of the 5-minute interval',
         y='Average number of steps taken', x='Interval')
```
### Which 5-minute interval contains the maximum number of steps?
```{r}
tdset2[which.max(tdset2$steps), ]$interval
```

## Imputing missing values
### total number of missing values in the dataset
```{r}
sum(!complete.cases(data))
```
### filling missing NAs by the previously calculated mean for the 5-minute interval
```{r}
tdset2$imputed_steps <- floor(tdset2$steps)
imputed_data <- merge(data, tdset2[,c('interval', 'imputed_steps')], by='interval')
imputed_data$steps <- ifelse(is.na(imputed_data$steps),
                                 imputed_data$imputed_steps,
                                 imputed_data$steps)
imputed_data$imputed_steps <- NULL
sum(is.na(imputed_data))
```
### mean number of steps taken per day after filling missing NAs
```{r}
melted_data3 <- melt(imputed_data, id.vars="date")
daily_imputed_data <- dcast(melted_data3, date ~ variable, sum)
mean(daily_imputed_data$steps)
```
### median number of steps taken per day after filling missing NAs
```{r}
median(daily_imputed_data$steps)
```
### Replace the data in the original histogram with the imputed data
```{r}
p %+% daily_imputed_data +
  labs(title='Number of steps taken each day,\nafter imputing missing values')
```

The mean does not differ much from the previous value but median does differ a little bit. So the impact of imputing missing data on the estimates of the total daily number of steps is negligibly small.

## Are there differences in activity patterns between weekdays and weekends?
### Creates a new factor variable with two levels – “weekday” and “weekend” 
```{r}
daysID <- function(date) {
    if (weekdays(as.Date(date)) %in% c("Saturday", "Sunday")  ) {
        "weekend"
    } 
    else {
        "weekday"
    }
}
imputed_data$daysID <- as.factor(sapply(imputed_data$date, daysID))
imputed_data$date <- NULL
table(imputed_data$daysID)
```
```{r}
### Panel plot containing a time series plot of the 5-minute interval 
par(mfrow = c(2, 1))
for (type in c("weekend", "weekday")) {
    steps.type <- aggregate(steps ~ interval, data = imputed_data, subset = imputed_data$daysID == 
        type, FUN = mean)
    plot(steps.type, type = "l", main = type)
}
```
